---
title: "Simulation Project"
author: "Morgane Flauder"
date: "December 27, 2015"
output: pdf_document
---


# The Central Limit Theorem and the exponential distribution

In this project, I will investigate the exponential distribution and compare it to the Central Limit Theorem. I will compare the sample mean and variance of simulations with the theoretical mean and variance of the exponential distribution. For this purpose, I will do 1000 simulations of the averages of 40 exponentials. I will set the parameter $\lambda$ to 0.2.

## Comparison between the sample mean and the theoretical mean

Since I used $\lambda = 0.2$, the theoretical mean is $\mu = \frac{1}{\lambda} = \frac{1}{0.2} = 5$.

I created a vector of 1000 simulations for the means of 40 exponentials.

```{r}
# Simulations of the means
set.seed(60)
mexp <- NULL
for (i in 1:1000) mexp <- c(mexp, mean(rexp(40,0.2)))
```

The mean of these 1000 simulations is :

```{r}
mean(mexp)
```

We can see that it's really close to the theoretical mean of 5.


## Comparison between the sample variance of the mean and the theoretical variance of the mean

We know that the variance of the sample mean is the population variance divided by $n$. The sample variance of the mean is $\frac{\sigma ^2}{n} = \frac{(1/\lambda)^2}{n} = \frac{(1/0.2)^2}{n} = \frac{25}{40} = 0.625$.

```{r}
var(mexp)
```

We can see that the sample variance of the mean is close to the theoretical variance of the mean of 0.625.

## Comparison between the sample variance and the theoratical variance

We know that the variance of the exponential distribution is $\sigma^2$. Since $\lambda = 0.2$ and $\sigma = \frac{1}{\lambda}$, the theoretical variance is $\sigma^2 = (\frac{1}{\lambda})^2 = (\frac{1}{0.2})^2 = 5^2 = 25$.

Let's do 1000 simulations of the variance of 40 exponentials.

```{r}
# Simulations of the variances
set.seed(60)
vexp <- NULL
for (i in 1:1000) vexp <- c(vexp, var(rexp(40,0.2)))
```

Let's compute the mean of the variances.

```{r}
mean(vexp)
```

We can see that this value is close to the theoretical variance of 25.


## Distribution

We are now going to check that the distribution of the means of the simulations follows a normal distribution.

We know that $\frac{\bar X_n - \mu}{\sigma / \sqrt{n}}$ follows a standard normal distribution.
To verify this, I computed $\frac{mexp - \mu}{\sigma / \sqrt{n}} = \frac{mexp - 5}{5 / \sqrt{40}}$.
I plotted the density of this vector and compared it to a standard normal distribution.

```{r}
sigma <- 5
mu <- 5
n <- 40
norm_mexp <- (mexp - mu) / (sigma / sqrt(n))

# Plot
hist(norm_mexp,
     freq = FALSE,
     breaks = 20,
     main = "Density of the averages of 1000 simulations\nof 40 exponential distributions",
     xlab = "x",
     ylim = c(0,0.4))
# Vertical line on the sample mean
abline(v = mean(norm_mexp), 
       col = "red", 
       lwd = 4)
# Normal distribution reference
curve(dnorm(x,0,1), 
      add = TRUE, 
      col = "darkblue", 
      lwd=2)
legend("topright",
       legend = c(paste("Sample Mean = ", round(mean(norm_mexp), digits = 2)), 
                  "Standard normal distribution"),
       col = c("red", "darkblue"),
       lty = c(1,1),
       lwd = 2,
       cex = 0.75)

```


We can see that the sample mean is centered on 0, the theoretical mean of the standard normal distribution, and follow roughly a standard normal distribution of $\mu = 0$ and $\sigma = 1$.











